
---
title: "Data621-Week04-Discussion"
author: "Pavan Akula"
date: "February 21, 2018"
output:
  html_document:
    code_folding: hide
---

I want to demonstrate how to identify outliers and high leverage and influential data points. I will be using `Advertisement Revenue` dataset from website http://www.stat.tamu.edu/~sheather/book/data_sets.php.

Plot the data using `boxplot`.

```{r, echo=T, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)
library(ggrepel)
library(knitr)    # Report display, table format
library(kableExtra)

AdRevDf <- read.csv("https://raw.githubusercontent.com/akulapa/Data621-Week04-Discussion/master/AdRevenue.csv", header= TRUE, stringsAsFactors = F)
AdRevDf$XY <- paste0("(",round(AdRevDf$Circulation,2), " ,", round(AdRevDf$AdRevenue,2),")")

p1 <- ggplot(data = AdRevDf, aes(x = "", y = Circulation)) + geom_boxplot() + labs(title="Boxplot: Circulation",x="") + coord_flip()
p2 <-ggplot(data = AdRevDf, aes(x = "", y = AdRevenue)) + geom_boxplot() + labs(title="Boxplot: Advertisement Revenue",x="") + coord_flip()

AdRevDf$BxPOutlierC <- ifelse(AdRevDf$Circulation>(summary(AdRevDf$Circulation)[5]+1.5*IQR(AdRevDf$Circulation)),'Yes','No')
AdRevDf$BxPOutlierA <- ifelse(AdRevDf$AdRevenue>(summary(AdRevDf$AdRevenue)[5]+1.5*IQR(AdRevDf$AdRevenue)),'Yes','No')
AdRevDf$Outlier <- ifelse((AdRevDf$BxPOutlierC == 'Yes'|AdRevDf$BxPOutlierA == 'Yes'),'Yes','No')

grid.arrange(p1, p2, nrow=2, newpage = F)
```

According to `boxplot` any observation that lies outside `lower limit` $(Q1 - 1.5\times IQR)$ and `upper limit` $(Q3 + 1.5\times IQR)$ is considered `outlier`.

####Outliers - Circulation

```{r, echo=T, message=F, warning=F}
summary(AdRevDf$Circulation)

AdRevDf %>% filter(BxPOutlierC == 'Yes') %>% arrange(desc(Circulation)) %>% select(Magazine, AdRevenue, Circulation) %>% 
  kable(format="html", caption = "Outliers Based on Circulation") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")

```

####Outliers - Advertisement Revenue

```{r, echo=T, message=F, warning=F}
summary(AdRevDf$AdRevenue)

AdRevDf %>% filter(BxPOutlierA == 'Yes') %>% arrange(desc(AdRevenue)) %>% select(Magazine, AdRevenue, Circulation) %>% 
  kable(format="html", caption = "Outliers Based on Advertisement Revenue") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")

```


####Regression Equation


```{r}

AdRevDf1 <- AdRevDf %>% filter(BxPOutlierA == 'No' & BxPOutlierC == 'No')

withOL<-lm(AdRevenue~Circulation, data=AdRevDf)
withoutOL<-lm(AdRevenue~Circulation, data=AdRevDf1)


z <- list(xx = format(coef(withOL)[1], digits = 4),
          yy = format(coef(withOL)[2], digits = 4),
          r2 = format(summary(withOL)$r.squared, digits = 3));

eq <- substitute(italic(hat(y)) == xx + yy %.% italic(x)*","~~italic(r)^2~"="~r2,z)

withOLeq <- as.character(as.expression(eq))

z <- list(xx = format(coef(withoutOL)[1], digits = 4),
          yy = format(coef(withoutOL)[2], digits = 4),
          r2 = format(summary(withoutOL)$r.squared, digits = 3));

eq <- substitute(italic(hat(y)) == xx + yy %.% italic(x)*","~~italic(r)^2~"="~r2,z)

withoutOLeq <- as.character(as.expression(eq))

ggplot(data=AdRevDf, aes(Circulation,AdRevenue)) + 
  geom_point(aes(col=Outlier)) + 
  scale_color_manual(values=c("black", "red")) +
  scale_y_continuous(limits=c(-50,900)) +
  geom_abline(intercept = coef(withOL)[1], slope = coef(withOL)[2], color="blue") +
  geom_abline(intercept = coef(withoutOL)[1], slope = coef(withoutOL)[2], color="purple") +
  geom_text_repel(data=filter(AdRevDf, (BxPOutlierC == 'Yes'|BxPOutlierA == 'Yes')), aes(label=XY), size=3) +
  annotate("text", x = 20, y = 400, label = withOLeq, colour="blue", size = 3, parse=T) +
  annotate("text", x = 20, y = 350, label = "(With Outliers)", colour="blue", size = 3) + 
  annotate("text", x = 10, y = 750, label = withoutOLeq, colour="purple", size = 3, parse=T) +
  annotate("text", x = 10, y = 700, label = "(Without Outliers)", colour="purple", size = 3)

```

Linear model summary with outliers.

```{r}
summary(withOL)
```

Linear model summary without outliers.

```{r}
summary(withoutOL)
```

Leverage values
```{r}
AdRevDf$Leverage <- hatvalues(withOL)

AdRevDfHat <- AdRevDf %>% arrange(desc(Leverage)) %>% slice(1:10) %>% select(Magazine, XY, Leverage, Outlier)

ggplot(AdRevDfHat, aes(seq_along(Leverage), Leverage)) + geom_col() + geom_text(data=AdRevDfHat, aes(label=round(Leverage,2)), size=3, vjust=0) + labs(x="", y="Leverage", title = "Top 10 Leverage Values")

```

```{r}
AdRevDfHat %>%  
  kable(format="html", caption = "Top 10 Leverage Values") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")
```

Observations with high leverage. 

If `leverage` value of an observation is greater than three times the mean leverage value ($3\times \frac{p}{n}$),  it is considered as high leverage observation. `p` is number of parameters (intercept $\beta_0$ and slope $\beta_1$) and `n` is number of observations.

Mean leverage value = $\frac{p}{n} = \frac{2}{70} = `r round(2/70,4) `$

Three times mean leverage value = `r 3 * round(2/70,4) `

```{r}
AdRevDfHat <- AdRevDf %>% filter(Leverage > 3 * round(2/70,4)) %>% select(Magazine, XY, Leverage, Outlier)

AdRevDfHat %>%  
  kable(format="html", caption = "High Leverage Values") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")

```


####Analysis

- The $R^2$ value has decreased from `r summary(withOL)$r.squared` to `r summary(withoutOL)$r.squared` when `outliers` are excluded from the model. We can conclude that relationship between `AdRevenue` and `Circulation` is moderately strong in the presence of `outliers`. Without the presence of `outliers` relationship is weak.

- The standard error of $\beta_1$ is almost two times more when `outliers` are removed. Value incresed from `r coef(withOL)[2]` to `r coef(withoutOL)[2]`. Higher value of $\beta_1$ leads to increase in width of confidence interval for $\beta_1$. Cofidence interval is calculated as $\bar x \pm z\times \frac{se}{\sqrt n}$. Where `se` is standard error of `Circulation` ($\beta_1$). 

- With and without outliers, linear model resulted in `p-value` for $H_0: \beta_1 = 0$ less than `0.0001`. We can say with 95% confidence that `AdRevenue` is function `Circulation` and both are related.

- In the presence of outliers predicted responses($\beta_0$) and estimated slope coefficient ($\beta_1$) are changing, we can conclude data points are highly influential.

- Based on `leverage` value we can conclude almost all observations that are identified by `boxplot` as `outliers` are high leverage data points.

In summary, data points are not `outliers`, but `highly influential leverage points`. Data points can be excluded from linear model.

####References

- https://onlinecourses.science.psu.edu/stat501/node/337
- http://www.whatissixsigma.net/box-plot-diagram-to-identify-outliers/
- https://www.rdocumentation.org/packages/VGAM/versions/1.0-4/topics/hatvalues
- http://ggplot2.tidyverse.org/reference/fortify.lm.html

